Index measures for oak decline severity using phenotypic descriptors

### 2019-12-12 manuscript/manuscript.Rmd
added to methods section

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  bookdown::word_document2:
   toc: false
bibliography: references.bib
---

@@

```{r loadData, include = FALSE}
loadd(site_rf_margin,
      site_rf_post_correction_margin,
      PDI_status_ttest,
      DAI_groups_ttest)
```

@@

R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and analysis of phenotypic data [@R2019].
The code used to generate the decline indexes is available as an R package pdi (v) at https://github.com/jasenfinch/pdi.
All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.

@@

Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius). 
An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.  
This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.
Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.

@@

- Decline index calculation

The randomForest R package

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest.
The proximity 
The relative proximities and the phenotypic descriptors multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:

@@

###

@@

@@

- LIME analysis to identify locally important descriptors under circumstances of interest
- Top ranked descriptor index response surfaces

To understand the descriptors contributing to each of the indexes, individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index. 
Crown related descriptors were found to contribute most to the PDI (Figure #) with more severely declining trees having reduced crown volume and condition. 
The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.

>>>>

---
title: "Index measures for oak decline severity using phenotypic descriptors"
output: 
  #bookdown::word_document2:
  # toc: false
  bookdown::pdf_document2: default
bibliography: references.bib
---

@@

```{r loadData, include = FALSE}
loadd(site_rf_margin,
      site_rf_post_correction_margin,
      PDI_status_ttest,
      DAI_groups_ttest,
      PDI_rf_tune_params,
      DAI_rf_tune_params)
```

@@

R v`r str_c(version$major,version$minor,sep ='.')` was used for all parsing, preparation and subsequent analysis of phenotypic data [@R2019].
The code used to generate the decline indexes is available as an R package pdi v`r packageVersion('pdi') %>% as.character()` at https://github.com/jasenfinch/pdi.
All raw data and code used to generate this manuscript are available at https://github.com/jasenfinch/Index_measures_for_oak_decline_severity_using_phenotypic_descriptors.

@@

To account for these differences and to make the trees across the sites comparable in the context of decline, an adjustment of site means to the overall mean was applied to each of the size descriptors (diameter at breast height, lower crown height, timber height, total height, crown radius). 
An example of this adjustment is shown for diameter at breast height in Supplementary Figure # and the adjustment factors for each of the descriptors at each site is shown in Supplementary Table #.  
This adjustment removed inter-site variability but maintained intra-site variability, allowing comparisons to be made between the trees at the different sites in the context of decline severity.

@@

The phenotypic descriptors were analysed using 100 repetitions of unsupervised random forest using the randomForest R package v`r packageVersion('randomForest') %>% as.character()` [@liaw2002].
The proximity 
The relative proximities and the phenotypic descriptors multidimensional scaling (MDS) to two dimensions was performed to yield two phenotypic indexes:

@@

### Overall and local descriptor importance

- LIME analysis to identify locally important descriptors under circumstances of interest
- simulated descriptor response surfaces

Individual random forest regression models of the phenotypic descriptors were trained, supervised using each decline index.
Optimal number of trees (ntree) and number randomly of variables (mtry) parameters for random forest were tuned prior to training. 
The combination parameters giving the lowest percent mean absolute error (PMAE) were selected and as follows: PDI - `r glue(ntree = {PDI_rf_tune_params$ntree}, mtry = {PDI_rf_tune_params$mtry}, PMAE = {PDI_rf_tune_params$PMAE %>% signif(digits = 3)})`; DAI - ntree = , mtry = , PMAE = (Supplementary figure #)
To understand the descriptors contributing to each of the indexes, Local Interpretable Model-Agnostic Explanations (LIME) using the lime R package v`r packageVersion('lime') %>% as.character()` [@ribeiro2016; @pedersen2019]

@@

Large phenotypic differences were found between the sites by supervised random forest (Supplementary  figure #a; margin = `r site_rf_margin$Margin %>% signif(digits = 3)`). 
This was as a results of differences in descriptors relating tree size and likely the ages of the trees between the sites. 
Post adjustment the random forest margin was substantially reduced to `r site_rf_post_correction_margin$Margin %>% signif(digits = 3)` indicating that the majority of the between site variance had been removed.

@@

- Accuracies of retrained models
- Lime results
- Top ranked descriptor index response surfaces

Crown related descriptors were found to contribute most to the PDI (Figure #) with more severely declining trees having reduced crown volume and condition. 
The DAI was a more complex in descriptor contribution with trunk staining and bleeding descriptors contributing highly with bleed prevalence contributing the most. 
Trees with high bleed prevalence and therefore a larger area of their trunk covered with black staining or active bleeds had much higher DAI scores.
